---
title: "Ortholog Analysis"
author: "Zoe Dellaert"
date: "2026-05-16"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Genome annotation

## Load packages

```{r load-packages}
library(tidyverse)
library(data.table)
library(igraph)

# set up some universal vectors
species <- c("Mcap","Pacuta","Pcomp")
```

## This code finds orthologs between protein sequences of the genomes generated by [Stephens et al. 2022 Publication](https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755?login=false) and various other coral genomes

1.  Run broccoli with the following genomes
    1.  Aiptasia (Cui et al., 2023; 10.1126/sciadv.adf7108)
    2.  Hydra (Siebert et al., 2019; 10.1126/science.aav9314)
    3.  Hydractinia (Salamanca-Díaz et al., 2025; 10.1038/s41467-025-57168-z)
    4.  Nematostella (Cole et al., 2024;10.1186/s12983-024-00529-z)
    5.  Astrangia (used to map Oculina data by Valadez-Ingersoll et al., 2024; 10.1101/2024.06.20.599951)
    6.  Orbicella (del Campo et al., 2024; 10.21203/rs.3.rs-5397639/v1)
    7.  Pocillopora damicornis (Diao et al., 2025; 10.3389/fmars.2025.1551763)
    8.  Stylophora (Levy et al., 2021; 10.1016/j.cell.2021.04.005)
    9.  Xenia (Hu et al., 2020; 10.1038/s41586-020-2385-7)
    10. ***Pocillopora acuta***
    11. ***Porites compressa***
    12. ***Montipora capitata***
    
### Step 1: Aquire genome protein fasta sequences

```{bash, eval=FALSE}
# from Rmd directory (HI_genome_annotations/code/)
mkdir -p ../references/cnidarian_protein_fastas
cd ../references/cnidarian_protein_fastas

# 1 - Aiptasia, assembled by Baumgarten et al., 2015 (10.1073/pnas.1513318112)
wget http://aiptasia.reefgenomics.org/download/aiptasia_genome.proteins.fa.gz

# 2 - Hydra, assembled by Steele et al 2015 (https://research.nhgri.nih.gov/hydra/)
wget https://research.nhgri.nih.gov/hydra/download/genemodels_proteins/hydra2.0_genemodels.aa.gz

# 3 - Hydractinia, assembled by Kon-Nanjo et al 2023 (10.1093/g3journal/jkad107)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/029/227/915/GCF_029227915.1_HSymV2.1/GCF_029227915.1_HSymV2.1_protein.faa.gz

# 4 - Nematostella, assembled by Zimmermann et al., 2023 (10.1038/s41467-023-44080-7)
wget https://simrbase.stowers.org/files/pub/nematostella/Nvec/genomes/Nvec200/aligned/tcs_v2/20240221/NV2g.20240221.protein.fa

# 5 - Astrangia, assembled by Stankiewicz et al., 2025 (10.1093/g3journal/jkaf033)
wget -O apoculata_proteins.fasta.gz https://zenodo.org/records/14110456/files/apoculata_proteins.fasta.gz?download=1

# 6 - Orbicella, assembled by Young et al., 2024 (10.1186/s12864-024-10092-w)
wget -O Orbicella_faveolata_gen_17.proteins.fa https://zenodo.org/records/10151798/files/Orbicella_faveolata_gen_17.proteins.fa?download=1

# 7 - Pocillopora damicornis, assembled by Cunning et al., 2018 (10.1038/s41598-018-34459-8)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/003/704/095/GCF_003704095.1_ASM370409v1/GCF_003704095.1_ASM370409v1_protein.faa.gz

# 8 - Stylophora, assembled by Voolstra et al., 2017 (10.1038/s41598-017-17484-x)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/571/385/GCF_002571385.2_Stylophora_pistillata_v1.1/GCF_002571385.2_Stylophora_pistillata_v1.1_protein.faa.gz

# 9 - Xenia, assembled by Hu et al., 2020 (10.1038/s41586-020-2385-7)
wget https://cmo.carnegiescience.edu/endosymbiosis/genome/xenSp1.proteins.fa

# Pocillopora acuta, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.pep.faa.gz

# Porites compressa, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/porites_compressa/Porites_compressa_HIv1.genes.pep.faa.gz

# Montipora capitata, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.pep.faa.gz

gunzip *

mv aiptasia_genome.proteins.fa Edia_proteins.faa
mv hydra2.0_genemodels.aa Hvul_proteins.faa
mv GCF_029227915.1_HSymV2.1_protein.faa Hsym_proteins.faa
mv NV2g.20240221.protein.fa Nvec_proteins.faa
mv apoculata_proteins.fasta Apoc_proteins.faa
mv Orbicella_faveolata_gen_17.proteins.fa Ofav_proteins.faa
mv GCF_003704095.1_ASM370409v1_protein.faa Pdam_proteins.faa
mv GCF_002571385.2_Stylophora_pistillata_v1.1_protein.faa Spis_proteins.faa
mv xenSp1.proteins.fa Xesp_proteins.faa
mv Pocillopora_acuta_HIv2.genes.pep.faa Pacu_proteins.faa
mv Porites_compressa_HIv1.genes.pep.faa Pcom_proteins.faa
mv Montipora_capitata_HIv3.genes.pep.faa Mcap_proteins.faa
```

for reference: *file_shortname*

```
Edia_proteins.faa
Hvul_proteins.faa
Hsym_proteins.faa
Nvec_proteins.faa
Apoc_proteins.faa
Ofav_proteins.faa
Pdam_proteins.faa
Spis_proteins.faa
Xesp_proteins.faa
Pacu_proteins.faa
Pcom_proteins.faa
Mcap_proteins.faa
```

### Step 2: Running broccoli on 12 cnidarian genomes

```{bash, eval = FALSE}
# from Rmd directory (HI_genome_annotations/code/)
nano 04_broccoli.sh
```

```{bash, eval = FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --ntasks=1 --cpus-per-task=20 #split one task over multiple CPU
#SBATCH --mem=500GB
#SBATCH -p cpu-preempt
#SBATCH -t 48:00:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error="%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output="%x_output.%j" #once your job is completed, any final job report comments will be put in this file

#load conda and activate conda environment
module load conda/latest
conda activate /work/pi_hputnam_uri_edu/conda/envs/env-broccoli

#load additional programs needed to run broccoli
module load uri/main
module load diamond/2.1.7
module load all/FastTree/2.1.11-GCCcore-12.3.0

mkdir -p ../references/cnidarian_orthologs/broccoli/

cd ../references/cnidarian_orthologs/broccoli/
python ~/broccoli.py -dir ../../cnidarian_protein_fastas/ -ext '.faa' -path_fasttree FastTree -threads 8
```

#### broccoli v2 - maximum likelihood analysis (takes longer, more accurate)

```{bash, eval = FALSE}
cd cnidarian_marker_genes/
nano scripts/broccoli_max_likelihood.sh
```

```{bash, eval = FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --ntasks=1 --cpus-per-task=24 #split one task over multiple CPU
#SBATCH --mem=500GB
#SBATCH -q long
#SBATCH -t 72:00:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error="%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output="%x_output.%j" #once your job is completed, any final job report comments will be put in this file

#load conda and activate conda environment
module load conda/latest
conda activate /work/pi_hputnam_uri_edu/conda/envs/env-broccoli

#load additional programs needed to run broccoli
module load uri/main
module load diamond/2.1.7
module load all/FastTree/2.1.11-GCCcore-12.3.0

mkdir -p ../references/cnidarian_orthologs/broccoli/max_likelihood

cd ../references/cnidarian_orthologs/broccoli/max_likelihood

python ~/broccoli.py -dir ../../cnidarian_protein_fastas/ -phylogenies 'ml' -ext '.faa' -path_fasttree FastTree -threads 24
```


### Step 3. Load in results from ortholog analyses

```{r}
# broc_pairs <- data.table::fread("../references/cnidarian_orthologs/broccoli/dir_step4/orthologous_pairs.txt", header=F)
# colnames(broc_pairs) <- c("Gene_spA","Gene_spB")

broc_pairs <- data.table::fread("../../snRNA_analysis/multi-sp-snRNA/reference_genes/cnidarian_marker_genes/output/broccoli/dir_step4/orthologous_pairs.txt", header=F)
colnames(broc_pairs) <- c("Gene_spA","Gene_spB")
```

### Step 4. Process to keep only these three species

```{r}
broc_pairs_3sp <- broc_pairs %>%
  # find rows that contain genes from our three species and label them with a new species column
  mutate(
    species_A = case_when(
      grepl("Pocillopora_acuta", Gene_spA) ~ "Pacuta",
      grepl("Montipora_capitata", Gene_spA) ~ "Mcap",
      grepl("Porites_compressa", Gene_spA) ~ "Pcomp"
    ),
    species_B = case_when(
      grepl("Pocillopora_acuta", Gene_spB) ~ "Pacuta",
      grepl("Montipora_capitata", Gene_spB) ~ "Mcap",
      grepl("Porites_compressa", Gene_spB) ~ "Pcomp"
    )
  ) %>%
  # remove rows that don't contain genes from our three species
  filter(!is.na(species_A), !is.na(species_B)) %>%
  select(Gene_spA, species_A, Gene_spB, species_B)
```

### Step 5. Construct high-confidence orthogroups from Broccoli Step 4 pairwise analysis 

In Broccoli step 3 orthogroup analysis, there are genes assigned to multiple orthogroups, which causes some downstream issues when trying to do cross-species comparisons. Here, I use the broc_pairs_3sp dataset to construct three-species orthogroups from these 1:1 pairs.

#### Build orthogroups from pairwise relationships

```{r}
# Create graph from pairwise relationships
edges <- broc_pairs_3sp %>%
  select(Gene_spA, Gene_spB)

g <- graph_from_data_frame(edges, directed = FALSE)
summary(g)

# Find components == orthogroups - for each gene it is assigned to one orthogroup/component only
orthogroups <- components(g)
```

```{r}
# Create orthogroup IDs
gene_to_og <- data.frame(
  # get the gene ids from all the genes in the pair list
  gene_id = names(orthogroups$membership),
  # add "OG_" label in front of the numbered orthogroups/compoenents
  OG = paste0("OG_", orthogroups$membership)) %>%
  # add species label column
  mutate(
    species = case_when(
      grepl("Pocillopora_acuta", gene_id) ~ "Pacuta",
      grepl("Montipora_capitata", gene_id) ~ "Mcap",
      grepl("Porites_compressa", gene_id) ~ "Pcomp"
    )
  )

cat("\nOrthogroups built from pairs:", n_distinct(gene_to_og$OG))
```

#### Get summary info for each orthogroup created

We want to know - which species have a gene in the OG, how many, and classify any orthogroups that have one gene in each species (OG_1to1to1)

```{r}
og_stats <- gene_to_og %>%
  group_by(OG) %>%
  summarise(
    n_total_genes = n(),
    n_species = n_distinct(species),
    n_Pacuta = sum(species == "Pacuta"),
    n_Mcap = sum(species == "Mcap"),
    n_Pcomp = sum(species == "Pcomp"),
    .groups = 'drop'
  ) %>%
  mutate(
    OG_in_all_3 = (n_Pacuta > 0) & (n_Mcap > 0) & (n_Pcomp > 0),
    OG_1to1to1 = (n_Pacuta == 1) & (n_Mcap == 1) & (n_Pcomp == 1)
  )

# Join stats back to gene list
ortho_pairs_final <- gene_to_og %>%
  left_join(og_stats, by = "OG")

## Validation: there should only be one row for each gene, and this also tells us how many genes per species are in the list
ortho_pairs_final %>%
  group_by(species, gene_id) %>%
  summarise(n_rows_per_gene = n(), .groups = 'drop') %>%
  group_by(species,n_rows_per_gene) %>%
  summarize(unique_genes=n())
```

#### Species-level summaries of orthogroups:

- n_genes = how many genes for this species are represented
- n_in_all_3 = of these genes, how many are in an orthogroup that has genes from all three species in it
- n_1to1to1 = of these genes, how many are in an orthogroup that has exactly one gene from all three species in it

```{r}
ortho_pairs_final %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_all_3 = sum(OG_in_all_3),
    pct_in_all_3 = 100*mean(OG_in_all_3),
    n_1to1to1 = sum(OG_1to1to1),
    pct_1to1to1 = 100*mean(OG_1to1to1)
  )
```

```{r}
write.csv(ortho_pairs_final, 
          "../annotation/orthologs_3sp.csv", 
          row.names = FALSE)

# Also save the pairwise relationships for reference
write.csv(broc_pairs_3sp, 
          "../annotation/ortholog_pairwise_relationships.csv", 
          row.names = FALSE)
```
