---
title: "Ortholog Analysis"
author: "Zoe Dellaert"
date: "2026-05-16"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Genome annotation

## Load packages

```{r load-packages}
library(tidyverse)
library(data.table)
library(igraph)

# set up some universal vectors
species <- c("Mcap","Pacuta","Pcomp")
```

## This code finds orthologs between protein sequences of the genomes generated by [Stephens et al. 2022 Publication](https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755?login=false) and various other coral genomes

1.  Run broccoli with the following genomes
    1.  Aiptasia (Cui et al., 2023; 10.1126/sciadv.adf7108)
    2.  Hydra (Siebert et al., 2019; 10.1126/science.aav9314)
    3.  Hydractinia (Salamanca-Díaz et al., 2025; 10.1038/s41467-025-57168-z)
    4.  Nematostella (Cole et al., 2024;10.1186/s12983-024-00529-z)
    5.  Astrangia (used to map Oculina data by Valadez-Ingersoll et al., 2024; 10.1101/2024.06.20.599951)
    6.  Orbicella (del Campo et al., 2024; 10.21203/rs.3.rs-5397639/v1)
    7.  Pocillopora damicornis (Diao et al., 2025; 10.3389/fmars.2025.1551763)
    8.  Stylophora (Levy et al., 2021; 10.1016/j.cell.2021.04.005)
    9.  Xenia (Hu et al., 2020; 10.1038/s41586-020-2385-7)
    10. ***Pocillopora acuta***
    11. ***Porites compressa***
    12. ***Montipora capitata***
    
### Step 1: Aquire genome protein fasta sequences

```{bash, eval=FALSE}
# from Rmd directory (HI_genome_annotations/code/)
mkdir -p ../references/cnidarian_protein_fastas
cd ../references/cnidarian_protein_fastas

# 1 - Aiptasia, assembled by Baumgarten et al., 2015 (10.1073/pnas.1513318112)
wget http://aiptasia.reefgenomics.org/download/aiptasia_genome.proteins.fa.gz

# 2 - Hydra, assembled by Steele et al 2015 (https://research.nhgri.nih.gov/hydra/)
wget https://research.nhgri.nih.gov/hydra/download/genemodels_proteins/hydra2.0_genemodels.aa.gz

# 3 - Hydractinia, assembled by Kon-Nanjo et al 2023 (10.1093/g3journal/jkad107)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/029/227/915/GCF_029227915.1_HSymV2.1/GCF_029227915.1_HSymV2.1_protein.faa.gz

# 4 - Nematostella, assembled by Zimmermann et al., 2023 (10.1038/s41467-023-44080-7)
wget https://simrbase.stowers.org/files/pub/nematostella/Nvec/genomes/Nvec200/aligned/tcs_v2/20240221/NV2g.20240221.protein.fa

# 5 - Astrangia, assembled by Stankiewicz et al., 2025 (10.1093/g3journal/jkaf033)
wget -O apoculata_proteins.fasta.gz https://zenodo.org/records/14110456/files/apoculata_proteins.fasta.gz?download=1

# 6 - Orbicella, assembled by Young et al., 2024 (10.1186/s12864-024-10092-w)
wget -O Orbicella_faveolata_gen_17.proteins.fa https://zenodo.org/records/10151798/files/Orbicella_faveolata_gen_17.proteins.fa?download=1

# 7 - Pocillopora damicornis, assembled by Cunning et al., 2018 (10.1038/s41598-018-34459-8)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/003/704/095/GCF_003704095.1_ASM370409v1/GCF_003704095.1_ASM370409v1_protein.faa.gz

# 8 - Stylophora, assembled by Voolstra et al., 2017 (10.1038/s41598-017-17484-x)
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/571/385/GCF_002571385.2_Stylophora_pistillata_v1.1/GCF_002571385.2_Stylophora_pistillata_v1.1_protein.faa.gz

# 9 - Xenia, assembled by Hu et al., 2020 (10.1038/s41586-020-2385-7)
wget https://cmo.carnegiescience.edu/endosymbiosis/genome/xenSp1.proteins.fa

# Pocillopora acuta, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.pep.faa.gz

# Porites compressa, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/porites_compressa/Porites_compressa_HIv1.genes.pep.faa.gz

# Montipora capitata, assembled by Stephens et al., 2022 (10.1093/gigascience/giac098)
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.pep.faa.gz

gunzip *

mv aiptasia_genome.proteins.fa Edia_proteins.faa
mv hydra2.0_genemodels.aa Hvul_proteins.faa
mv GCF_029227915.1_HSymV2.1_protein.faa Hsym_proteins.faa
mv NV2g.20240221.protein.fa Nvec_proteins.faa
mv apoculata_proteins.fasta Apoc_proteins.faa
mv Orbicella_faveolata_gen_17.proteins.fa Ofav_proteins.faa
mv GCF_003704095.1_ASM370409v1_protein.faa Pdam_proteins.faa
mv GCF_002571385.2_Stylophora_pistillata_v1.1_protein.faa Spis_proteins.faa
mv xenSp1.proteins.fa Xesp_proteins.faa
mv Pocillopora_acuta_HIv2.genes.pep.faa Pacu_proteins.faa
mv Porites_compressa_HIv1.genes.pep.faa Pcom_proteins.faa
mv Montipora_capitata_HIv3.genes.pep.faa Mcap_proteins.faa
```

for reference: *file_shortname*

```
Edia_proteins.faa
Hvul_proteins.faa
Hsym_proteins.faa
Nvec_proteins.faa
Apoc_proteins.faa
Ofav_proteins.faa
Pdam_proteins.faa
Spis_proteins.faa
Xesp_proteins.faa
Pacu_proteins.faa
Pcom_proteins.faa
Mcap_proteins.faa
```

### Step 2: Running broccoli on 12 cnidarian genomes

```{bash, eval = FALSE}
# from Rmd directory (HI_genome_annotations/code/)
nano 04_broccoli.sh
```

```{bash, eval = FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --ntasks=1 --cpus-per-task=20 #split one task over multiple CPU
#SBATCH --mem=500GB
#SBATCH -p cpu-preempt
#SBATCH -t 48:00:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error="%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output="%x_output.%j" #once your job is completed, any final job report comments will be put in this file

#load conda and activate conda environment
module load conda/latest
conda activate /work/pi_hputnam_uri_edu/conda/envs/env-broccoli

#load additional programs needed to run broccoli
module load uri/main
module load diamond/2.1.7
module load all/FastTree/2.1.11-GCCcore-12.3.0

mkdir -p ../references/cnidarian_orthologs/broccoli/

cd ../references/cnidarian_orthologs/broccoli/
python ~/broccoli.py -dir ../../cnidarian_protein_fastas/ -ext '.faa' -path_fasttree FastTree -threads 8
```

#### broccoli v2 - maximum likelihood analysis (takes longer, more accurate)

```{bash, eval = FALSE}
cd cnidarian_marker_genes/
nano scripts/broccoli_max_likelihood.sh
```

```{bash, eval = FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --ntasks=1 --cpus-per-task=24 #split one task over multiple CPU
#SBATCH --mem=500GB
#SBATCH -q long
#SBATCH -t 72:00:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error="%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output="%x_output.%j" #once your job is completed, any final job report comments will be put in this file

#load conda and activate conda environment
module load conda/latest
conda activate /work/pi_hputnam_uri_edu/conda/envs/env-broccoli

#load additional programs needed to run broccoli
module load uri/main
module load diamond/2.1.7
module load all/FastTree/2.1.11-GCCcore-12.3.0

mkdir -p ../references/cnidarian_orthologs/broccoli/max_likelihood

cd ../references/cnidarian_orthologs/broccoli/max_likelihood

python ~/broccoli.py -dir ../../cnidarian_protein_fastas/ -phylogenies 'ml' -ext '.faa' -path_fasttree FastTree -threads 24
```


### Step 3. Load in results from ortholog analyses

```{r}
broc_pairs <- data.table::fread("../references/cnidarian_orthologs/broccoli/dir_step4/orthologous_pairs.txt", header=F)
colnames(broc_pairs) <- c("Gene_spA","Gene_spB")
```

### Step 4. Process to keep only these three species

```{r}
broc_pairs_3sp <- broc_pairs %>%
  # find rows that contain genes from our three species and label them with a new species column
  mutate(
    species_A = case_when(
      grepl("Pocillopora_acuta", Gene_spA) ~ "Pacuta",
      grepl("Montipora_capitata", Gene_spA) ~ "Mcap",
      grepl("Porites_compressa", Gene_spA) ~ "Pcomp"
    ),
    species_B = case_when(
      grepl("Pocillopora_acuta", Gene_spB) ~ "Pacuta",
      grepl("Montipora_capitata", Gene_spB) ~ "Mcap",
      grepl("Porites_compressa", Gene_spB) ~ "Pcomp"
    )
  ) %>%
  # remove rows that don't contain genes from our three species
  filter(!is.na(species_A), !is.na(species_B)) %>%
  select(Gene_spA, species_A, Gene_spB, species_B)
```

### Step 5. Construct high-confidence orthogroups from Broccoli Step 4 pairwise analysis 

In Broccoli step 3 orthogroup analysis, there are genes assigned to multiple orthogroups, which causes some downstream issues when trying to do cross-species comparisons. Here, I use the broc_pairs_3sp dataset to construct three-species orthogroups from these 1:1 pairs.

#### Build orthogroups from pairwise relationships

```{r}
# Create graph from pairwise relationships
edges <- broc_pairs_3sp %>%
  select(Gene_spA, Gene_spB)

g <- graph_from_data_frame(edges, directed = FALSE)
summary(g)

# Find components == orthogroups - for each gene it is assigned to one orthogroup/component only
orthogroups <- components(g)
```

```{r}
# Create orthogroup IDs
gene_to_og <- data.frame(
  # get the gene ids from all the genes in the pair list
  gene_id = names(orthogroups$membership),
  # add "OG_" label in front of the numbered orthogroups/compoenents
  OG = paste0("OG_", orthogroups$membership)) %>%
  # add species label column
  mutate(
    species = case_when(
      grepl("Pocillopora_acuta", gene_id) ~ "Pacuta",
      grepl("Montipora_capitata", gene_id) ~ "Mcap",
      grepl("Porites_compressa", gene_id) ~ "Pcomp"
    )
  )

cat("\nOrthogroups built from pairs:", n_distinct(gene_to_og$OG))
```

#### Get summary info for each orthogroup created

We want to know - which species have a gene in the OG, how many, and classify any orthogroups that have one gene in each species (OG_1to1to1)

```{r}
og_stats <- gene_to_og %>%
  group_by(OG) %>%
  summarise(
    n_total_genes = n(),
    n_species = n_distinct(species),
    n_Pacuta = sum(species == "Pacuta"),
    n_Mcap = sum(species == "Mcap"),
    n_Pcomp = sum(species == "Pcomp"),
    .groups = 'drop'
  ) %>%
  mutate(
    OG_in_all_3 = (n_Pacuta > 0) & (n_Mcap > 0) & (n_Pcomp > 0),
    OG_1to1to1 = (n_Pacuta == 1) & (n_Mcap == 1) & (n_Pcomp == 1)
  )

# Join stats back to gene list
ortho_pairs_final <- gene_to_og %>%
  left_join(og_stats, by = "OG")

## Validation: there should only be one row for each gene, and this also tells us how many genes per species are in the list
ortho_pairs_final %>%
  group_by(species, gene_id) %>%
  summarise(n_rows_per_gene = n(), .groups = 'drop') %>%
  group_by(species,n_rows_per_gene) %>%
  summarize(unique_genes=n())
```

#### Species-level summaries of orthogroups:

- n_genes = how many genes for this species are represented
- n_in_all_3 = of these genes, how many are in an orthogroup that has genes from all three species in it
- n_1to1to1 = of these genes, how many are in an orthogroup that has exactly one gene from all three species in it

```{r}
ortho_pairs_final %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_all_3 = sum(OG_in_all_3),
    pct_in_all_3 = 100*mean(OG_in_all_3),
    n_1to1to1 = sum(OG_1to1to1),
    pct_1to1to1 = 100*mean(OG_1to1to1)
  )
```

```{r}
write.csv(ortho_pairs_final, 
          "../annotation/orthologs_3sp.csv", 
          row.names = FALSE)

# Also save the pairwise relationships for reference
write.csv(broc_pairs_3sp, 
          "../annotation/ortholog_pairwise_relationships.csv", 
          row.names = FALSE)
```

<!-- ### Biomineralization Toolkit -->

<!-- Biomineralization Toolkit gene list from [Scucchia et al. 2021](https://onlinelibrary.wiley.com/doi/full/10.1111/gcb.15812). -->

<!-- #### Biomineralization list based on BLAST analysis -->

<!-- - reformatted fasta file provided by F. Scucchia so that each header began on a new line, so that each sequence is on one single line, and to remove duplicate genes or genes not in the accompanying excel file. -->
<!-- - There are 172 genes in the "toolkit"  -->

<!-- ```{bash eval=FALSE} -->
<!-- #request interactive computing node -->
<!-- salloc -p cpu -c 8 --mem 32G -->

<!-- #load BLAST module -->
<!-- module load uri/main -->
<!-- module load BLAST+/2.15.0-gompi-2023a -->
<!-- cd input_list/biomineralization -->

<!-- # run blast with tabular output -->
<!-- blastp -query Biomineralization_Toolkit_FScucchia_ZDrefmt.fasta -db ../../../references/blast_dbs/Pacuta_prot -out Biomineralization_blast_results_tab.txt -outfmt 6 -evalue 0.01 -max_target_seqs 1 -->
<!-- ``` -->

<!-- Now, will take the best Pacuta alignment for each Biomineralization Gene and match to the name of that gene -->

<!-- ```{r} -->
<!-- Biomin_genes <- read_xlsx("input_lists/biomineralization/Biomineralization_Toolkit_FScucchia.xlsx") %>% select(-`blasted protein in Stylophora`) -->
<!-- Biomin_blast_results <- read.delim("input_lists/biomineralization/Biomineralization_blast_results_tab.txt", header=FALSE)  -->
<!-- Biomin_blast_results <- Biomin_blast_results %>% filter(V11 < 0.01) %>% select(V1, V2) %>% distinct() -->

<!-- # Merge data frames based on accessionnumber/geneID -->
<!-- merged_data <- Biomin_genes %>% -->
<!--   inner_join(Biomin_blast_results, by = c("accessionnumber/geneID" = "V1")) %>% dplyr::rename("Pocillopora_acuta_best_hit" = "V2") %>% mutate(List="Biomin_Genes") -->

<!-- write.csv(merged_data, "genes_of_interest/Pacuta_Biomin_Blast.csv", row.names = F) -->
<!-- ``` -->

<!-- #### Biomineralization list based on ortholog analysis -->

<!-- ```{r} -->
<!-- Biomin_Spis <- read_xlsx("input_lists/biomineralization/Biomineralization_Toolkit_FScucchia.xlsx")  %>% clean_names() -->

<!-- Biomin_Spis <- Biomin_Spis %>%  separate_longer_delim(blasted_protein_in_stylophora, delim=",") %>%  -->
<!--       mutate(blasted_protein_in_stylophora=str_squish(blasted_protein_in_stylophora))  -->
<!-- length(unique(Biomin_Spis$blasted_protein_in_stylophora)) -->

<!-- Biomin_broc <- Biomin_Spis %>%  -->
<!--     left_join(broc_pairs_clean, by = c("blasted_protein_in_stylophora"="Gene_spA")) %>% -->
<!--     bind_rows( -->
<!--       Biomin_Spis %>% left_join(broc_pairs_clean, by = c("blasted_protein_in_stylophora"="Gene_spB")) %>% -->
<!--         dplyr::rename(Gene_spB = Gene_spA)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- broc_groups <- data.table::fread("cnidarian_marker_genes/output/broccoli/dir_step3/table_OGs_protein_names.txt", header=T) -->

<!-- broc_groups_Spis <- broc_groups %>% select(c(`#OG_name`,Spis_proteins.faa))  %>% filter(Spis_proteins.faa != "") %>% dplyr::rename(OG_name = "#OG_name") -->
<!-- broc_groups_Spis_sep <- broc_groups_Spis %>% separate_longer_delim(cols = Spis_proteins.faa, delim = " ") -->
<!-- broc_groups_Spis_sep <- broc_groups_Spis_sep %>% dplyr::rename(query = Spis_proteins.faa) -->

<!-- broc_groups_Pacu <- broc_groups %>% select(c(`#OG_name`,Pacu_proteins.faa))  %>% filter(Pacu_proteins.faa != "") %>% dplyr::rename(OG_name = "#OG_name") -->
<!-- broc_groups_Pacu_sep <- broc_groups_Pacu %>% separate_longer_delim(cols = Pacu_proteins.faa, delim = " ") -->
<!-- broc_groups_Pacu_sep <- broc_groups_Pacu_sep %>% dplyr::rename(query = Pacu_proteins.faa) -->

<!-- Biomin_broc_groups <- Biomin_Spis %>%  -->
<!--     left_join(broc_groups_Spis_sep, by = c("blasted_protein_in_stylophora"="query")) -->

<!-- Biomin_broc_groups_Pacu <- Biomin_broc_groups %>%  -->
<!--     left_join(broc_groups_Pacu_sep, by = "OG_name") -->

<!-- Biomin_broc_Pacuta <- Biomin_broc_groups_Pacu %>% filter(grepl("acuta",query)) %>% select(-blasted_protein_in_stylophora) %>% distinct() %>% mutate(List="Biomin_Genes") -->

<!-- write.csv(Biomin_broc_Pacuta, file="genes_of_interest/Pacuta_Biomin_Spis_orthogroups.csv") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- name <- "Spis" -->

<!-- print(paste0(name," number of unique gene ids in gene of interest list:")) -->
<!-- print(length(unique(Biomin_Spis$blasted_protein_in_stylophora))) -->

<!-- print(paste0(name," matches of gene id to broccoli protein names:")) -->
<!-- print(length(unique(union(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spA),intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spB))))) -->

<!-- print(paste0(name," matches of gene id to *unclassified* broccoli protein names:")) -->
<!-- print(length(unique(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_unclassified$V1)))) -->

<!-- print(paste0(name," unnacounted for (in broc groups only?):")) -->
<!-- length(unique(Biomin_Spis$blasted_protein_in_stylophora)) - -->
<!--   (length(unique(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_unclassified$V1))) + length(unique(union(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spA),intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spB))))) -->

<!-- setdiff(unique(Biomin_Spis$blasted_protein_in_stylophora), -->
<!--         union( -->
<!--           unique(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_unclassified$V1)), -->
<!--           unique(union(intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spA), -->
<!--                        intersect(Biomin_Spis$blasted_protein_in_stylophora,broc_pairs_clean$Gene_spB))))) -->

<!-- # one protein that has actually been removed from the Spis genome on ncbi due to contamination: XP_022803894.1 -->
<!-- ``` -->


<!-- ```{r} -->
<!-- Biomin_broc_Pacuta_temp <- Biomin_broc %>% filter(grepl("acuta",Gene_spB)) %>% dplyr::rename(Pacuta_gene=Gene_spB) -->

<!-- length(unique(Biomin_broc_Pacuta_temp$blasted_protein_in_stylophora)) -->
<!-- #98 of the 119 proteins with broccoli orthologs had Pacuta orthologs -->

<!-- # which ones did not: -->

<!-- Biomin_Spis %>% filter(blasted_protein_in_stylophora %in% setdiff(unique(Biomin_Spis$blasted_protein_in_stylophora),unique(Biomin_broc_Pacuta_temp$blasted_protein_in_stylophora))) -->

<!-- Biomin_Spis %>% filter(blasted_protein_in_stylophora %in% setdiff(unique(Biomin_Spis$blasted_protein_in_stylophora),unique(Biomin_broc_Pacuta_temp$blasted_protein_in_stylophora))) %>% pull(definition) %>% unique() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- Biomin_broc_Pacuta <- Biomin_broc %>% filter(grepl("acuta",Gene_spB)) %>% dplyr::rename(Pacuta_gene=Gene_spB) %>% select(-blasted_protein_in_stylophora) %>% distinct() %>% mutate(List="Biomin_Genes") -->

<!-- write.csv(Biomin_broc_Pacuta, file="genes_of_interest/Pacuta_Biomin_Spis_ortholog.csv") -->

<!-- Biomin_broc_Pdam<- Biomin_broc %>% filter(grepl("XP_027",Gene_spB)) %>% dplyr::rename(Pdam_gene=Gene_spB) %>% select(-blasted_protein_in_stylophora) %>% distinct() %>% mutate(List="Biomin_Genes") -->

<!-- write.csv(Biomin_broc_Pdam, file="genes_of_interest/Pdam_Biomin_Spis_ortholog.csv") -->

<!-- Biomin_broc_Pcomp <- Biomin_broc %>% filter(grepl("compressa",Gene_spB)) %>% dplyr::rename(Pcomp_gene=Gene_spB) %>% select(-blasted_protein_in_stylophora) %>% distinct() -->

<!-- write.csv(Biomin_broc_Pcomp, file="genes_of_interest/Pcomp_Biomin_Spis_ortholog.csv") -->

<!-- Biomin_broc_Mcap <- Biomin_broc %>% filter(grepl("capitata",Gene_spB)) %>% dplyr::rename(Mcap_gene=Gene_spB) %>% select(-blasted_protein_in_stylophora) %>% distinct() -->

<!-- write.csv(Biomin_broc_Mcap, file="genes_of_interest/Mcap_Biomin_Spis_ortholog.csv") -->
<!-- ``` -->

<!-- ### Heat Stress Genes -->

<!-- ```{r} -->
<!-- # load in manually curated list of genes of interest -->
<!-- stress_genes_man <- read.csv("input_lists/HeatStressGenes.csv", header = TRUE) -->

<!-- NEED_TO_FIND_GENE_ID <- stress_genes_man %>% filter(is.na(species_gene_ID)) -->
<!-- print(NEED_TO_FIND_GENE_ID) -->

<!-- stress_genes_man <- stress_genes_man %>% filter(!is.na(species_gene_ID)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- stress_genes <- stress_genes_man %>% separate_rows(species_gene_ID,sep =";") %>% filter(!is.na(species_gene_ID)) -->

<!-- unique(stress_genes$species) -->

<!-- ## clean gene IDs for Spis, only need to add .1 to the end of all the species_gene_ID - this is the format of the broc IDs -->
<!-- stress_genes_Spis <- stress_genes %>% filter(species=="Spis")  %>%  -->
<!--   mutate(protein_id = paste0(species_gene_ID,".1"))  -->

<!-- ## make protein_id column for the Edia + Pdam genes, which already have their IDs in the right format -->
<!-- stress_genes_Pdam <- stress_genes %>% filter(species=="Pdam") %>%  -->
<!--   mutate(protein_id = species_gene_ID) -->

<!-- stress_genes_Edia <- stress_genes %>% filter(species=="Edia")  %>%  -->
<!--   mutate(protein_id = species_gene_ID)  -->

<!-- ## combine and merge with broccoli orthologs -->
<!-- stress_genes_final <- rbind(stress_genes_Pdam,stress_genes_Spis,stress_genes_Edia) -->

<!-- stress_broc <- stress_genes_final %>%  -->
<!--     left_join(broc_pairs_clean, by = c("protein_id"="Gene_spA")) %>% -->
<!--     bind_rows( -->
<!--       stress_genes_final %>% left_join(broc_pairs_clean, by = c("protein_id"="Gene_spB")) %>% -->
<!--         dplyr::rename(Gene_spB = Gene_spA)) -->
<!-- ``` -->

<!-- Species-specific lists for species: -->

<!-- ```{r} -->
<!-- # Pocillopora acuta -->
<!-- stress_broc_Pacuta <- stress_broc %>% filter(grepl("acuta",Gene_spB)) %>% dplyr::rename(Pacuta_gene=Gene_spB) %>% -->
<!--   select(-c(species_gene_ID,protein_id))  %>%  #remove ID columns from the original species for clarity -->
<!--   distinct() %>% mutate(List="HeatStressGenes") -->
<!-- # Porites compressa  -->
<!-- stress_broc_Pcomp <- stress_broc %>% filter(grepl("compressa",Gene_spB)) %>% dplyr::rename(Pcomp_gene=Gene_spB) %>% -->
<!--   select(-c(species_gene_ID,protein_id))  %>%  #remove ID columns from the original species for clarity -->
<!--   distinct() %>% mutate(List="HeatStressGenes") -->
<!-- # Montipora capitata -->
<!-- stress_broc_Mcap <- stress_broc %>% filter(grepl("capitata",Gene_spB)) %>% dplyr::rename(Mcap_gene=Gene_spB) %>% -->
<!--   select(-c(species_gene_ID,protein_id))  %>%  #remove ID columns from the original species for clarity -->
<!--   distinct() %>% mutate(List="HeatStressGenes") -->
<!-- # Orbicella faveolata -->
<!-- stress_broc_Ofav <- stress_broc %>% filter(grepl("QW917",Gene_spB)) %>% dplyr::rename(Ofav_gene=Gene_spB) %>% -->
<!--   select(-c(species_gene_ID,protein_id))  %>%  #remove ID columns from the original species for clarity -->
<!--   distinct() %>% mutate(List="HeatStressGenes") -->

<!-- write.csv(stress_broc_Pacuta, file="genes_of_interest/HeatStressGenes_Pacuta.csv") -->
<!-- write.csv(stress_broc_Pcomp, file="genes_of_interest/HeatStressGenes_Pcomp.csv") -->
<!-- write.csv(stress_broc_Mcap, file="genes_of_interest/HeatStressGenes_Mcap.csv") -->
<!-- write.csv(stress_broc_Ofav, file="genes_of_interest/HeatStressGenes_Ofav.csv") -->
<!-- ``` -->

<!-- What are we missing? -->
<!-- ```{r} -->
<!-- NEED_TO_FIND_GENE_ID$gene_id -->

<!-- stress_genes_man$gene_id[!stress_genes_man$gene_id %in% stress_broc_Pacuta$gene_id] -->
<!-- stress_genes_man$gene_id[!stress_genes_man$gene_id %in% stress_broc_Pcomp$gene_id] -->
<!-- stress_genes_man$gene_id[!stress_genes_man$gene_id %in% stress_broc_Mcap$gene_id] -->
<!-- stress_genes_man$gene_id[!stress_genes_man$gene_id %in% stress_broc_Ofav$gene_id] -->

<!-- stress_genes_man[!stress_genes_man$gene_id %in% stress_broc_Pacuta$gene_id,] -->
<!-- stress_genes_man[!stress_genes_man$gene_id %in% stress_broc_Pcomp$gene_id,] -->
<!-- stress_genes_man[!stress_genes_man$gene_id %in% stress_broc_Mcap$gene_id,] -->
<!-- stress_genes_man[!stress_genes_man$gene_id %in% stress_broc_Ofav$gene_id,] -->
<!-- ``` -->

<!-- ## Appendix: Program installations -->

<!-- ### Install Broccoli on unity -->

<!-- Set up conda environments and install program -->

<!-- ```{bash, eval=FALSE} -->
<!-- # enter interactive session -->
<!-- salloc -->

<!-- # load conda -->
<!-- module load conda/latest -->

<!-- # create and activate environment -->
<!-- conda create --prefix /work/pi_hputnam_uri_edu/conda/envs/env-broccoli python=3.8 ete3 -->
<!-- conda activate /work/pi_hputnam_uri_edu/conda/envs/env-broccoli -->

<!-- # clone broccoli repo into environment -->
<!-- cd /work/pi_hputnam_uri_edu/conda/envs/env-broccoli -->
<!-- git clone https://github.com/rderelle/Broccoli.git -->
<!-- cd Broccoli -->

<!-- # check that the install is sucessful by running the help command -->
<!-- python broccoli.py -help -->

<!-- # Make a symlink (shortcut) to home directory -->
<!-- ln -s /work/pi_hputnam_uri_edu/conda/envs/env-broccoli/Broccoli/broccoli.py ~/broccoli.py -->

<!-- # now it can run in any directory via -->
<!-- python ~/broccoli.py -help -->
<!-- ``` -->
